Modal Verification Patterns for Systems

### 2025-06-23 paper.tex
comments

  \todo[inline]{Colin says: how about this framing instead of the second paragraph above:}
  \begin{comment}
  In this paper we examine how modalities have emerged as a common structure in formal verification
  of low-level software, and explain how many recent examples naturally share 
  common structure in the relationship between the modalities and software features they are used
  to reason about.
  We explain how the concept of a \emph{resource context} (a class of system resources
  to reason about) naturally corresponds to families of modal operators indexed by
  system data, and how this naturally leads to using modal assertions to describe
  \emph{resource elements} (data in the relevant context).
  We also describe extensions of this idea to ongoing work in formal verification
  of filesystems code.
  \end{comment}
\end{abstract}

@@

\section{Introduction}
Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.
  \todo[inline]{this section is already nice for introducing the ideas and what you mean by "systems software."
  This section mostly needs a little more background on systems concepts like virtual memory, or what it means
  to virtualize locations of blocks on disk}
Systems software, in general, interfaces with an underlying computing architecture such that any software system at any higher level in the software stack can (at least indirectly) utilize the resources of the machine.
The last layer of software before the hardware is naturally critical to the correctness of an overall system,
as essentially all software built on top of it assumes its correctness.
%Assuming verified hardware, this makes the correctness of the system software critical to the rest of the software stack. 
%Because they interface with the hardware, abstractions and their implementations are intricate and error-prone. 
And because hardware is complex and highly diverse, the implementation of those lowest layers
of the software stack is typically intricate and naturally error-prone, despite how critical its correctness is.
Typically systems software has, as a primary focus, the task of \emph{abstracting} from hardware details to simplify 
the construction of higher layers of the stack.
\subsection{Hybrid Logic Dynamic Logic and Nominals in Program Logics}
\label{sec:modalbackground}

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\loc \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. We adapt these ideas to assertions of the form $@[l]\varphi$asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

An example utilization of state naming explicitly on the assertion appears in program logics such as Iris~\cite{jung2018iris}, which enables encoding of usage protocols (e.g. state transition system) resembling typestate~\cite{strom1986typestate,garcia2014foundations} as specifications. Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
An example more familiar to the systems community is Halpern et al.'s history of adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

@@

\subsection{Virtualization in Systems Software}
\label{sec:background}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.
\paragraph{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
\todo{change the text}
Memory addresses seen by programs running on modern computers are \emph{virtualized}: the address observed by a running program generally will not correspond directly to the physical location in memory, and may not even correspond to a physical location that \emph{exists} in the machine. Instead, these \emph{virtual} addresses are translated to \emph{physical} addresses that correspond directly to locations in RAM. On most modern architectures, this translation is performed through cooperation of the hardware and OS kernel: while executing an instruction that dereferences a (virtual) address, the CPU's \emph{memory management unit} (\textsc{MMU}) implements an architecture-specific process of \emph{address translation}, resulting in a physical address used to access the cache\footnote{Technically, for performance reasons most caches are indexed with parts of the virtual address, but tagged with the physical data addresses, so cache lookups and address translations can proceed in parallel.} and/or memory-bus.

@@

On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables calle d \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} shows, address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table. On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were fewer levels. We omit those names to avoid confusion, and for greater consistency with the recent addition of a 5th level on the most recent processors, where levels 4 and 5 are simply known by their numbers.}  Because address translation traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ \texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.

@@

\begin{figure}
    \includegraphics[width=0.75\columnwidth]{cow_index.png}
    \caption{A Filesystem called \textsf{Modfs} Indexing its Physical Pages for File Pages with CoW (copy-on-write) Tree}
    \label{fig:cowindex}
\end{figure}

@@

\paragraph{Different Models in Ensuring Disk Consistency of Filesystems}
With their intricate semantics and critical place in modern systems, formal verification of functional correctness for storage systems (including filesystems and other kinds of storage) have drawn much attention~\cite{ntzik-gardner-15,ntzik2018concurrent,chen-crash-16}\todo{lots more cites}. Unlike our \textsc{CoW} filesystem \textsf{ModFs} which allocates new chunks and rewrites all the nodes from leaf to root on the write-amplification path, some filesystems allows in-place updates to handle the heavy workloads more efficiently in principle. Unlike \textsc{CoW} filesystems, filesystems with in-place updates need to implement some form of \emph{journalling} to ensure that a crash partway through an in-place update can be rolled back (using an undo journal that records the original contents before modification as abstarcted in Figure \ref{fig:journal}) or completed (using a write-ahead journal that first writes new data in unused space before \emph{also} updating in place).\footnote{\textsf{CoW} filesystems can be viewed as a twist on write-ahead journals that uses the journal as the actual data once written to a fresh location, thereby reclaiming the original location. This results in a performance and complexity trade-off: journalling writes twice for every updated block and requires recovery logic to roll back or re-execute the log, while \textsf{CoW} techniques write each update only once and require slightly more intricate allocation strategies to compact fragmentation.}

Essentially, successfully writing the new superblock for a new snapshot is analagous to completing a valid journal entry, except it \emph{also} has the effect of updating the live filesystem data due to making the ``journal entries'' valid filesystem data and metadata blocks (an idea that goes back to early work on log-structured filesystems~\cite{Rosenblum1992LFS}).
In fact, we can think of the atomicity of the recovery from the crash state conceptually equal to the single block disk write persisting the change from $\textsf{Super}_0$ to $\textsf{Super}_1$ in Figure \ref{fig:cow_index}, therefore, considering \textsc{CoW} semantics, we do not need to consider the specification and proof of crash condition separately; instead a global never-violated invariant (not even temporarily broken) of the filesystem ensures the on-disk state is always consistent.
\paragraph{Memory Reclamation in Client Code} In a typical configuration, allocating, deallocating and referencing memory locations for a client code leave the memory layout in a form of DAG (directed-asyclic-graph) with a unique root which resembles the first stack pointer referencing the allocated memory. To ease picturing the memory layout after a set of requests from the client code (for allocation, referencing etc.), we can think of the unique roots of DAGs as \emph{super} block addresses -- this time multiple ones, not only two -- in Figure \ref{fig:cow_index}, from which a path is created for a stack pointer reference to the allocated memory. To reclaim the allocated memory location that is no longer being accessed, systems uses a well-known method called \emph{reference-counting} in which a new reference to the allocated memory increases the reference counter on it w.r.t a certain regime. 
\paragraph{Consistency of Reclaimed Memory} A sound memory reclamation is based on the conditions where the unique is not \emph{shared}, and there is no \emph{reachability} from the root to any the references pointing to the allocated memory. Once the requirements are satisfied then the memory location can be reclaimed \emph{soundly} -- i.e., without violating any access through a reference to it.
\paragraph{Buffering} Buffering is a technique used in variety of systems (filesystem page caching, write buffer in multi-processor architectures, translation-lookaside-buffers for translated addresses etc.) to delay the costly operation (committing file pages, flushing buffered writes from write-cache to memory etc.) making updates consistent.
\paragraph{Consistency of Reads of Buffered Writes} Consistency of reads of any value potentially buffered depends on the semantics of the operation making them \emph{observable}. A typical architecture such as Intel-x86 considers the \emph{flush} of buffered writes \emph{non-deterministically} -- i.e., at any time at any number of buffered writes. On the other hand, a typical filesystem can follow a certain policy to commit the in-memory file pages to the disk.   
\subsection{Conclusion \& Contributions}
\paragraph{Conclusion} Low-level systems exhibit certain patterns in their designs, especially when interacting with computing resources. Exploiting certain patterns while designing software has been an important field of study. In this regard, we think that certain properties of modalities enable us to understand and do the verification challenges that show certain patterns.

\paragraph{Contributions} We argue how modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\begin{itemize}
\item \textbf{Identifying System Verification Challenges}: We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\item \textbf{Introducing the Concept of \emph{Resource}}: Then we discuss the concept of \emph{resource} which has already been an essential concept in the design of systems. Inspired by the concept of resource in the systems, we define what a resource and its context are in the modal abstractions.
\item \textbf{Introducing the Concept of \emph{Nominals}}:
\todo[inline]{Colin says: I'd reframe the way nominals are introduced. This discussion is good,
but is a "systems-first" explanation. At least some of this audience will already know
what classic nominals are, so need to be shown naming specific resources as an application
of what they already know. (And some readers will need a very short intro/refresher.)
}
Nominalization enables identifying a resource in a context. For example, a transaction is a context of resources of in-memory updated disk blocks. The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, in case of a crash while persisting updated disk-blocks, the filesystem can rollback the \emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. To be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction identifier -- \emph{strong nominalization}. In another example, virtual memory references (resources) in an address space (resource context), which can be uniquely identified with a root address (the nominal) of its page-table tree, are \emph{agnostic} to the address space that they are in. However, they can only be accessed (be valid) in the address space to which they are agnostic. However, then an address space switch happens, the virtual memory references of the previous address space must be made inaccessible. To be able to do so, although virtual memory references do not hold any piece of information related to their address space, we still have to associate them. We call this kind of unilateral nominalization -- \emph{weak nominalization}.
\item \textbf{Taxonomy of the Current Modal Approaches in System Verification}: Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, weak memory verification, storage persistence ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2024modalabstractionsvirtualizingmemory,amalreal2024} because we would like to justify that our \emph{definitions} are not domain dependent.
\end{itemize}
%\section{Resources in Systems Software}
%\label{sec:systemsoft}

@@

To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again,casting them in a certain way to bring out the relevance of our philosophy.

\todo[inline]{Colin says: I don't think this section needs to change much,
but it should be read carefully with the new audience in mind,
possibly giving more details on what systems concepts each system is dealing with, for a non-systems audience.}
\paragraph{Shallow Modalities in Iris Separation Logic}
\label{sec:modal_in_iris}
Classically modal operators are given direct semantics in terms of an underlying model, which for \textsf{Iris} would be a collection of \textsc{CMRA}s; this is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{krebbers2017essence}. However in general, many modalities can be given proof-theoretic semantics directly within a logic. This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} logic itself), as well as for a wide range of other modalities in the literature~\cite{restall2002introduction}. These modalities are typically multi-modal, with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ is typically replaced with substructural implication $\sepimp$.

>>>>

  \todo[inline]{Colin says: how about this framing instead of the second paragraph above:}
  \begin{comment}
  In this paper we examine how modalities have emerged as a common structure in formal verification
  of low-level software, and explain how many recent examples naturally share 
  common structure in the relationship between the modalities and software features they are used
  to reason about.
  We explain how the concept of a \emph{resource context} (a class of system resources
  to reason about) naturally corresponds to families of modal operators indexed by
  system data, and how this naturally leads to using modal assertions to describe
  \emph{resource elements} (data in the relevant context).
  We also describe extensions of this idea to ongoing work in formal verification
  of filesystems code.
  \end{comment}
  \todo[inline,color=cyan]{Ismail, why did you comment out my suggestion and keep the original?}
\end{abstract}

@@

\section{Introduction}
Although they differ in the functionality they offer, low-level systems exhibit certain patterns of design and utilization of computing resources.
Systems software, in general, interfaces with an underlying computing substrate such that any software system at any higher level in the software stack can (at least indirectly) utilize the resources of the machine.
The last layer of software before the hardware is naturally critical to the correctness of an overall system,
as essentially all software built on top of it assumes its correctness.
%Assuming verified hardware, this makes the correctness of the system software critical to the rest of the software stack. 
%Because they interface with the hardware, abstractions and their implementations are intricate and error-prone. 
And because hardware is complex and highly diverse, the implementation of those lowest layers
of the software stack is typically intricate and naturally error-prone, despite how critical its correctness is.
Typically systems software has, as a primary focus, the task of \emph{abstracting} from hardware details to simplify 
the construction of higher layers of the stack.

% \subsection{Conclusion \& Contributions}
% \paragraph{Conclusion} 
Low-level systems software exhibits certain patterns in their designs, especially when interacting with computing resources. 
Exploiting certain patterns while designing software has been an important field of study.
This survey examines these common patterns in systems software, and their relationships
to modalities. We believe that certain properties of modalities enable us to understand and 
address the verification challenges of these systems by tailoring specification and proof
to the same patterns used to design the systems.

\paragraph{Contributions} We argue that modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\begin{itemize}
\item \textbf{Identifying System Verification Challenges}: We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\item \textbf{Introducing the Concept of \emph{Resource}}: Then we discuss the concept of \emph{resource} which has already been an essential concept in the design of systems. Inspired by the concept of resource in the systems, we define what a resource and its context are in the modal abstractions.
\item \textbf{Introducing the Concept of \emph{Nominals for System Resources}}:
Nominalization enables identifying a resource in a context. 
For example, a transaction is a context of resources of in-memory updated disk blocks. i
The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, 
in case of a crash while persisting updated disk-blocks, the filesystem can rollback the 
\emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. 
To be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction 
identifier -- \emph{strong nominalization}. In another example, virtual memory references (resources) in 
an address space (resource context), which can be uniquely identified with a root address (the nominal) 
of its page-table tree, are \emph{agnostic} to the address space that they are in. 
However, they can only be accessed (be valid) in the address space to which they are agnostic. 
However, then an address space switch happens, the virtual memory references of the previous address 
space must be made inaccessible. To be able to do so, although virtual memory references do not hold any 
piece of information related to their address space, we still have to associate them. 
We call this kind of unilateral nominalization -- \emph{weak nominalization}.
\todo[inline,color=cyan]{I left this paragraph alone, mostly, but it's too detailed for the new intro section position here}
\item \textbf{Taxonomy of Current Modal Approaches in System Verification}: Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, reasoning about weak memory models and storage persistence ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2024modalabstractionsvirtualizingmemory,amalreal2024} because we would like to justify that our \emph{definitions} are not domain dependent.
\end{itemize}

\section{Background on Modal Logics for Programs, and Low-Level Systems Software}

This section briefly recalls related concepts from modal logic of particular relevance
to our taxonomy, which should be familiar but benefit from fixing notation (Section \ref{sec:modalbackground}),
and enough general background on the systems software concepts organized by our taxonomy
for the connections to modalities to be clear (Section \ref{sec:background}).

\subsection{Hybrid Logic Dynamic Logic and Nominals in Program Logics}
\label{sec:modalbackground}

@@

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\l \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.\todo{seems out of place?}
In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

An example utilization of state naming explicitly on the assertion appears in program logics such as Iris~\cite{jung2018iris}, which enables encoding of usage protocols (e.g. state transition system) resembling typestate~\cite{strom1986typestate,garcia2014foundations} as specifications. Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
One classic application of this idea to systems is 
Halpern et al.'s work adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

@@

\subsection{Virtualization in Systems Software}
\label{sec:background}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.
\paragraph{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
\todo{change the text}
Memory addresses seen by programs running on modern computers are \emph{virtualized}: the address observed by a running program generally will not correspond directly to the physical location in memory, and may not even correspond to a physical location that \emph{exists} in the machine. Instead, these \emph{virtual} addresses are translated to \emph{physical} addresses that correspond directly to locations in RAM. On most modern architectures, this translation is performed through cooperation of the hardware and OS kernel: while executing an instruction that dereferences a (virtual) address, the CPU's \emph{memory management unit} (\textsc{MMU}) implements an architecture-specific process of \emph{address translation}, resulting in a physical address used to access the cache\footnote{Technically, for performance reasons most caches are indexed with parts of the virtual address, but tagged with the physical data addresses, so cache lookups and address translations can proceed in parallel.} and/or memory-bus.

@@

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}
On the x86-64 architecture, the input to the \textsc{MMU}'s address translation is a sparse hierarchical set of tables called \emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} shows, address translation proceeds by repeatedly taking designated slices of the virtual address and indexing into a table. On x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.\footnote{Technically levels 1--3 have designated names originating when there were fewer levels. We omit those names to avoid confusion, and for greater consistency with the recent addition of a 5th level on the most recent processors, where levels 4 and 5 are simply known by their numbers.}  Because address translation traverses a series of tables, the translation process or algorithm is sometimes referred to as a \emph{page-table walk}. While Figure \ref{fig:pagetables} and most of our details are specific to the x86-64 architecture, ARMv8 (a.k.a.\ \texttt{aarch64}), RISC-V, and PowerPC use similar hierarchical page tables for address translation.

@@

\begin{figure}[t]
    \includegraphics[width=0.75\columnwidth]{cow_index.png}
    \caption{A Filesystem called \textsf{Modfs} Indexing its Physical Pages for File Pages with CoW (copy-on-write) Tree}
    \label{fig:cowindex}
\end{figure}

@@

\paragraph{Different Models in Ensuring Disk Consistency of Filesystems}
\todo[inline,color=cyan]{This looks like it's verbatim copy-pasted from the ModFS draft; if this exact text
it used in this publication, it means we'd need to completely rewrite the relevant text in that draft.}
With their intricate semantics and critical place in modern systems, formal verification of functional correctness for storage systems (including filesystems and other kinds of storage) have drawn much attention~\cite{ntzik-gardner-15,ntzik2018concurrent,chen-crash-16}\todo{lots more cites}. Unlike our \textsc{CoW} filesystem \textsf{ModFs} which allocates new chunks and rewrites all the nodes from leaf to root on the write-amplification path, some filesystems allows in-place updates to handle the heavy workloads more efficiently in principle. Unlike \textsc{CoW} filesystems, filesystems with in-place updates need to implement some form of \emph{journalling} to ensure that a crash partway through an in-place update can be rolled back (using an undo journal that records the original contents before modification as abstarcted in Figure \ref{fig:journal}) or completed (using a write-ahead journal that first writes new data in unused space before \emph{also} updating in place).\footnote{\textsf{CoW} filesystems can be viewed as a twist on write-ahead journals that uses the journal as the actual data once written to a fresh location, thereby reclaiming the original location. This results in a performance and complexity trade-off: journalling writes twice for every updated block and requires recovery logic to roll back or re-execute the log, while \textsf{CoW} techniques write each update only once and require slightly more intricate allocation strategies to compact fragmentation.}

Essentially, successfully writing the new superblock for a new snapshot is analagous to completing a valid journal entry, except it \emph{also} has the effect of updating the live filesystem data due to making the ``journal entries'' valid filesystem data and metadata blocks (an idea that goes back to early work on log-structured filesystems~\cite{Rosenblum1992LFS}).
In fact, we can think of the atomicity of the recovery from the crash state conceptually equal to the single block disk write persisting the change from $\textsf{Super}_0$ to $\textsf{Super}_1$ in Figure \ref{fig:cow_index}, therefore, considering \textsc{CoW} semantics, we do not need to consider the specification and proof of crash condition separately; instead a global never-violated invariant (not even temporarily broken) of the filesystem ensures the on-disk state is always consistent.
\paragraph{Memory Reclamation in Client Code} In a typical configuration, allocating, deallocating and referencing memory locations for a client code leave the memory layout in a form of DAG (directed-asyclic-graph) with a unique root which resembles the first stack pointer referencing the allocated memory. To ease picturing the memory layout after a set of requests from the client code (for allocation, referencing etc.), we can think of the unique roots of DAGs as \emph{super} block addresses -- this time multiple ones, not only two -- in Figure \ref{fig:cow_index}, from which a path is created for a stack pointer reference to the allocated memory. To reclaim the allocated memory location that is no longer being accessed, systems uses a well-known method called \emph{reference-counting} in which a new reference to the allocated memory increases the reference counter on it w.r.t a certain regime. 
\paragraph{Consistency of Reclaimed Memory} A sound memory reclamation is based on the conditions where the unique is not \emph{shared}, and there is no \emph{reachability} from the root to any the references pointing to the allocated memory. Once the requirements are satisfied then the memory location can be reclaimed \emph{soundly} -- i.e., without violating any access through a reference to it.
\paragraph{Buffering} Buffering is a technique used in variety of systems (filesystem page caching, write buffer in multi-processor architectures, translation-lookaside-buffers for translated addresses etc.) to delay the costly operation (committing file pages, flushing buffered writes from write-cache to memory etc.) making updates consistent.
\paragraph{Consistency of Reads of Buffered Writes} Consistency of reads of any value potentially buffered depends on the semantics of the operation making them \emph{observable}. A typical architecture such as Intel-x86 considers the \emph{flush} of buffered writes \emph{non-deterministically} -- i.e., at any time at any number of buffered writes. On the other hand, a typical filesystem can follow a certain policy to commit the in-memory file pages to the disk.

%\section{Resources in Systems Software}
%\label{sec:systemsoft}

@@

To explain our ideas in the general systems understanding, we briefly recap some of the background and themes our ideas build on again, casting them in a certain way to bring out the relevance of our philosophy.

\todo[inline]{Colin says: I don't think this section needs to change much,
but it should be read carefully with the new audience in mind,
possibly giving more details on what systems concepts each system is dealing with, for a non-systems audience.}
\paragraph{Shallow Modalities in Iris Separation Logic}
\label{sec:modal_in_iris}
\todo[inline,color=cyan]{this also seems taken from the VMM work}
Classically modal operators are given direct semantics in terms of an underlying model, which for \textsf{Iris} would be a collection of \textsc{CMRA}s; this is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{krebbers2017essence}. However in general, many modalities can be given proof-theoretic semantics directly within a logic. This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} logic itself), as well as for a wide range of other modalities in the literature~\cite{restall2002introduction}. These modalities are typically multi-modal, with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ is typically replaced with substructural implication $\sepimp$.
